### 2023-04-19 Manuscript/Draft1.Rmd
manuscipt updates

The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios).

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output. General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors and standard deviations. On the contrary, we also see individual cases where the stratified mean produced the lowest absolute error. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are included on sampling, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata). These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.

@@

Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each strata within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28). See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.

@@

The top two panels in Figure \@ref(fig:PopPct) depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences. While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

@@

<div align="center">Table 1. Parameters used in all population models.</div>

@@

tabl <- "
|	Parameter	|	Description | Unit |Yellowtail | Cod | Haddock | Source
|:---	|:------- |:-- |:--- |:-- |:-- |:--
| $\\rho$ | Ford's growth coefficient | wk$^{-1}$  | 4.48 | 4.43 | 4.49 | |
| $M$ | Natural Mortality| wk$^{-1}$  | 0.2064 | 0.2728 | 0.3340 | |
| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | |
| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | |
| $\\sigma^2$ | Variance in recruited fish | kg$^2$ | 0.55 | 0.55 | 0.55 ||
| $\\lambda$ | Decay rate for movement | - | 0.7 | 0.7 | 0.7 ||
| $Spwn_s$ | Spawning weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 ||
| $Rec_s$ | Recruitment weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 ||
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion

>>>>

The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \@ref(tab:scenarios). See Figure \@ref(fig:strata-plot) for a spatial reference of the Georges Bank strata and Figure \@ref(fig:hab-plot) for the static habitat preferences for each species ($Hab^2_{J,s}$) used in our population models.

Figure \@ref(fig:PopPct) depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28). The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. In these scenarios we notice a small amount of shifting between successive years as the population aggregates on hot spots in the domain. On the other hand, in the right column where the temperature is increasing over time we note significant shifting taking places in certain strata. This figure shows how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output. While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used. As a result, although $132/160=82.5\%$ of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

@@

The top two panels in Figure \@ref(fig:PopPct) depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, wihch is why we seen the same dynamics with constant and increasing temperature. While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are sampled, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata). Each of these instances are These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.

@@

<div align="center">Table 1. Parameters used in all population models. SAW 1: [@nefsc1218], SAW 2: [@nefsc1311], SAW 3: ??? </div>

@@

tabl <- "
|	Parameter	|	Description | Unit |Yellowtail | Cod | Haddock | Source
|:---	|:------- |:-- |:--- |:-- |:-- |:--
| $\\rho$ | Ford's growth coefficient | wk$^{-1}$  | 4.48 | 4.43 | 4.49 | [@thorson2020predicting] |
| $M$ | Natural Mortality| wk$^{-1}$  | 0.2064 | 0.2728 | 0.3340 | [@thorson2020predicting]|
| $F$ | Fishing Mortality| wk$^{-1}$  | 0.358 | 0.511 | 0.45 | SAW 1, 2, 3 |
| $W_R$ |Weight of fully recruited fish| kg | 0.39 | 2.95 | 1.12 | SAW 1, 2, 3 |
| $W_{R-1}$ | Weight of pre-recruit fish| kg | 0.13 | 0.39 | 0.19 | SAW 1, 2, 3 |
| $\\sigma^2$ | Variance in recruited fish | kg$^2$ | 0.55 | 0.55 | 0.55 |assumed|
| $\\lambda$ | Decay rate for movement | - | 0.7 | 0.7 | 0.7 |assumed|
| $Spwn_s$ | Spawning weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 |SAW 1, 2, 3|
| $Rec_s$ | Recruitment weeks for species $s$ | wk | 9-12 | 8-13 | 11-14 |SAW 1, 2, 3|
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion

### 2023-04-19 Manuscript/Draft1.tex
manuscipt updates

The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \ref{tab:scenarios}.

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error between abundance estimates and model output. General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors and standard deviations. On the contrary, we also see individual cases where the stratified mean produced the lowest absolute error. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are included on sampling, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}. These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.

@@

Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each strata within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28). See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.

@@

The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences. While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

@@

@@

@@

Table 1. Parameters used in all population models.

\begin{longtable}[]{@{}lllllll@{}}
\toprule
Parameter & Description & Unit & Yellowtail & Cod & Haddock & Source \\
\midrule
\endhead
\(\rho\) & Ford's growth coefficient & wk\(^{-1}\) & 4.48 & 4.43 & 4.49 & \\
\(M\) & Natural Mortality & wk\(^{-1}\) & 0.2064 & 0.2728 & 0.3340 & \\
\(W_R\) & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & \\
\(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & \\
\(\sigma^2\) & Variance in recruited fish & kg\(^2\) & 0.55 & 0.55 & 0.55 & \\
\(\lambda\) & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & \\
\(Spwn_s\) & Spawning weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & \\
\(Rec_s\) & Recruitment weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & \\
\bottomrule
\end{longtable}

>>>>

The goal of our project was to analyze the ability of contemporary population indexing methods to track population trends under a variety of conditions, as depicted in Table \ref{tab:scenarios}. See Figure \ref{fig:strata-plot} for a spatial reference of the Georges Bank strata and Figure \ref{fig:hab-plot} for the static habitat preferences for each species (\(Hab^2_{J,s}\)) used in our population models.

Figure \ref{fig:PopPct} depicts the spatial shifting that occurs in each stratum within our population models, specifically during the bottom trawl survey weeks in the spring (13 and 14) and fall (37 and 28). The left column depicts the percent of population that exists in each stratum for each species when the temperature is a constant repeating pattern. In these scenarios we notice a small amount of shifting between successive years as the population aggregates on hot spots in the domain. On the other hand, in the right column where the temperature is increasing over time we note significant shifting taking places in certain strata. This figure shows how we were able to mimic shifting distributions due to climate change by running simulations with a temperature gradient that increased on average over time.

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error between abundance estimates and model output. While the model-based results tended to provide a lower relative error to the stratified mean with a mean relative error of 0.34 compared to 0.38, the model-based estimates were very sensitive to the settings used. As a result, although \(132/160=82.5\%\) of individual scenarios had a VAST estimate with a lower relative error than the corresponding stratified mean estimate, the variance for the VAST errors was much larger than the stratified mean (0.10 compared to 0.02). VAST models that include covariate information provided the lowest overall errors and standard deviations. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

@@

@@

The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring dynamics are therefore related to the static habitat values in our model rather than the dynamics temperature preferences, wihch is why we seen the same dynamics with constant and increasing temperature. While we observe similar dynamics in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are sampled, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}. Each of these instances are These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.

@@

\leavevmode\vadjust pre{\hypertarget{ref-nefsc1218}{}}%
NEFSC. 2012. {54th Northeast Regional Stock Assessment Workshop (54th SAW) Assessment Report}. {Northeast Fish. Sci. Cent. Ref. Doc. 12-18; 604 p. Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.

\leavevmode\vadjust pre{\hypertarget{ref-nefsc1311}{}}%
NEFSC. 2013. {55th Northeast Regional Stock Assessment Workshop (55th SAW) Assessment Report}. {Northeast Fish. Sci. Cent. Ref. Doc. 13-11; 849 p. Available from: National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543-1026, or online at http://www.nefsc.noaa.gov/nefsc/publications/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-thorson2020predicting}{}}%
Thorson, J.T. 2020. Predicting recruitment density dependence and intrinsic growth rate for all fishes worldwide using a data-integrated life-history model. Fish and Fisheries \textbf{21}(2): 237--251. Wiley Online Library.

@@

Table 1. Parameters used in all population models. SAW 1: (NEFSC 2012), SAW 2: (NEFSC 2013), SAW 3: ???

\begin{longtable}[]{@{}lllllll@{}}
\toprule
Parameter & Description & Unit & Yellowtail & Cod & Haddock & Source \\
\midrule
\endhead
\(\rho\) & Ford's growth coefficient & wk\(^{-1}\) & 4.48 & 4.43 & 4.49 & (Thorson 2020) \\
\(M\) & Natural Mortality & wk\(^{-1}\) & 0.2064 & 0.2728 & 0.3340 & (Thorson 2020) \\
\(F\) & Fishing Mortality & wk\(^{-1}\) & 0.358 & 0.511 & 0.45 & SAW 1, 2, 3 \\
\(W_R\) & Weight of fully recruited fish & kg & 0.39 & 2.95 & 1.12 & SAW 1, 2, 3 \\
\(W_{R-1}\) & Weight of pre-recruit fish & kg & 0.13 & 0.39 & 0.19 & SAW 1, 2, 3 \\
\(\sigma^2\) & Variance in recruited fish & kg\(^2\) & 0.55 & 0.55 & 0.55 & assumed \\
\(\lambda\) & Decay rate for movement & - & 0.7 & 0.7 & 0.7 & assumed \\
\(Spwn_s\) & Spawning weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & SAW 1, 2, 3 \\
\(Rec_s\) & Recruitment weeks for species \(s\) & wk & 9-12 & 8-13 & 11-14 & SAW 1, 2, 3 \\
\bottomrule
\end{longtable}
